Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-18 index.qmd
section:introduction; process:write

  Dichotomous Comparative Judgment (DCJ, @Pollitt_2012a, @Pollitt_2012b) requires judges to evaluate the relative manifestation of traits between pairs of stimuli, resulting in a dichotomous outcome indicating which stimulus exhibits the trait more strongly. Research has demonstrated DCJ's effectiveness and reliability in various domains [@Pollitt_2012b; @Bartholomew_et_al_2018; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Boonen_et_al_2020]. However, the literature lacks a clear and transparent depiction of the plausible mechanisms underlying DCJ data. Specifically, there is no detail explanation of how the different assessment factors can potentially influence the observed DCJ data. This study aims to fill this gap by applying the framework of causal analysis and Directed Acyclic Graphs (DAG; @Pearl_2009). Using this framework, the study will construct a scientific model to elucidate the causal assumptions and mechanisms inherent the system. This model will enable researchers to draw inferences about causal relationships from DCJ data. Subsequently, the study will translate this model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference. The outcomes of this study will inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.

@@

<!-- 1. What is comparative judgment? -->
In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *Dichotomous* Comparative Judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b]. In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli. This assessment results in a dichotomous value indicating which stimulus possesses a higher degree of the trait. After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced. This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959; @Pollitt_2012a], which produces a latent variable corresponding to the trait of interest. This latent variable is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by the works of @Pollitt_2012b, @Bartholomew_et_al_2018, @vanDaal_et_al_2016, @Lesterhuis_2018, and @Bartholomew_et_al_2020. However, its application transcends education, as exemplified by the work of @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that give rise to DCJ data. Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data.

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and contribute to the reliability of the DCJ method. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. An example of this can be found in @vanDaal_et_al_2016, where the authors assessed the *skills in academic writing* (trait) of Flemish university students, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).

>>>>

  Dichotomous Comparative Judgment [DCJ, @Pollitt_2012a, @Pollitt_2012b] requires judges to evaluate the relative manifestation of traits between pairs of stimuli, resulting in a dichotomous outcome indicating which stimulus exhibits the trait more strongly. Research has demonstrated DCJ's effectiveness and reliability in various domains [@Pollitt_2012b; @Bartholomew_et_al_2018; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Boonen_et_al_2020]. Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that generate the DCJ data. Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data. This study aims to fill this gap by applying the framework of causal analysis and Directed Acyclic Graphs [DAG; @Pearl_2009]. Using this framework, the study will construct a scientific model to elucidate the causal assumptions and mechanisms inherent the system. This model will enable researchers to draw inferences about causal relationships from DCJ data. Subsequently, the study will translate this model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference. The outcomes of this study will inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.

@@

<!-- 1. What is comparative judgment? -->
In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *Dichotomous* Comparative Judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b]. In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli. This assessment results in a dichotomous value indicating which stimulus possesses a higher degree of the trait. After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced. This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959], which creates a score that corresponds with the trait of interest. This score is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by @Pollitt_2012b, @Jones_2015, @vanDaal_et_al_2016, @Bartholomew_et_al_2018, @Lesterhuis_2018, @Bartholomew_et_al_2020, and @Marshall_et_al_2020. However, its application transcends education, as exemplified by @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, there is no clear representation in the literature of the plausible mechanisms that generate DCJ data. Particularly, there is no depiction of the complexity and the underlying assumptions of the DCJ system, nor how different assessment factors can potentially influence the observed DCJ outcome.

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and influence the DCJ outcome. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. For instance, @vanDaal_et_al_2016 assessed university students' skills in academic writing, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).

<!-- 4. Solution to the gap -->
Although several studies have examined the individual impact of these factors on the method's reliability, including @Bramley_2015, @Pollitt_2012b, @Bramley_et_al_2019, @Verhavert_et_al_2019, @Crompvoets_et_al_2022, @vanDaal_et_al_2017, and @Gijsen_et_al_2021, to the best of the authors' knowledge, none have provided such transparent depiction of DCJ system and the plausible mechanisms that generate the DCJ outcome. This study aims to fill this gap by utilizing the framework of causal analysis and Directed Acyclic Graphs [DAG; @Pearl_2009; @Pearl_et_al_2016]. Using this framework, the study will construct a scientific model to elucidate the underlying assumptions of the DCJ system, providing plausible mechanisms of how the DCJ data is generated. This model will enable researchers to draw inferences about plausible causal relationships within the DCJ system. Furthermore, using a minimal set of assumptions from the framework, the study will translate the scientific model into a probabilistic statistical model, aiming to derive statistical estimands for different targets of inference. Ultimately, the results of this study could inform the planning of DCJ experiments and hold significance for researchers or analysts involved in education and assessment procedures who implement the DCJ methodology.
